A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-11 tex/example.tex
[ tex ] Revisions here and there

\section{Examples}\label{sec:example}
Since our theory is not confined to a specific simply-typed language, we can employ it with any language that can be specified by a bidirectional binding signature.
To showcase its general applicability, we present two additional examples:
\begin{enumerate}
\item \PCF with products and let bindings
\item spine calculus: a simply-typed $\lambda$-calculus with applications in spine form.
\end{enumerate}
The first is a standard exercise in implementing a type synthesiser and the second is an example of a system with infinitely many operations.

\subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This also applies to parsing: while there are grammar and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022} which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding following rules:
\bgroup
\small
  \begin{mathpar}
    \inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}
    \and
    \inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}
    \and
    \inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}
    \and
    \inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}
    \\
    \inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, x \chk A}
    \and
    \inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}
  \end{mathpar}
\egroup

In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
In this language, we introduce additional type constructs apart from $\bto$.
To do so, we extend $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with the operations $\mathsf{nat}$ and $\mathsf{times}$ such that
\[
  \arity({\mathsf{nat}}) = 0 \quad\text{and}\quad \arity({\mathsf{times}}) = 2
\]
enabling the introduction of $\mathtt{nat}$ and $A \times B$ as $\tyOp_{\mathsf{nat}}$ and $\tyOp_{\mathsf{times}}(A, B)$, respectively.
Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with an operation for each rule above:
\begin{flalign*}
  &  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}} \\
  &  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
  && \aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}
  && \aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\syn}} \\
  &  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
  && \aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}
\end{flalign*}
With these operations, a straightforward check shows that every rule is mode-correct, therefore this as a mode-correct bidirectional type system.
As a result, we can apply the type synthesiser presented in \Cref{sec:type-synthesis} directly to this system, eliminating the need for re-implementation.

\subsection{Spine Calculus}\label{subsec:spine}
A spine application, denoted as $t\;u_1\;\ldots\;u_n$, is a variant of application that consists of a head term $t$ and an indefinite number of arguments $u_1\;u_2\;\dots\;u_n$.
This arrangement allows direct access to the head term, making it practical in various applications.
For instance, \Agda's core language employs this form of applications, as does its reflected syntax used for metaprogramming.

At first glance, accommodating this form of application may seem impossible, given the number of arguments a construct can accept is finite as specified by a signature.
Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish corresponding constructs for each number $n$ of arguments, thereby exhibiting the necessity of having an arbitrary set (with decidable equality) for operation symbols.

>>>>

\section{Examples}\label{sec:example}
Our theory works for any simply typed language that can be specified in a syntax-directed manner by a bidirectional binding signature, and is not confined to a specific language.
To showcase its general applicability, we present two additional examples:
\begin{enumerate}
\item \PCF with products and let bindings, and
\item spine calculus, a simply typed $\lambda$-calculus with applications in spine form.
\end{enumerate}
The first is a standard exercise in implementing a type synthesiser and the second is an example of a system with infinitely many operations.

\subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This has also been the case for parsing: while there are grammar and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022}, which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding the following rules:
\bgroup
\small
  \begin{mathpar}
    \inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}
    \and
    \inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}
    \and
    \inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}
    \and
    \inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}
    \\
    \inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, x \chk A}
    \and
    \inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}
  \end{mathpar}
\egroup

%In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
In this language, we introduce additional type constructs apart from $\bto$.
To do so, we extend $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with the operations $\mathsf{nat}$ and $\mathsf{times}$ such that
\[
  \arity({\mathsf{nat}}) = 0 \quad\text{and}\quad \arity({\mathsf{times}}) = 2
\]
enabling the introduction of $\mathtt{nat}$ and $A \times B$ as $\tyOp_{\mathsf{nat}}$ and $\tyOp_{\mathsf{times}}(A, B)$, respectively.
Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with an operation for each rule above:
\begin{flalign*}
  &  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}} \\
  &  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
  && \aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}
  && \aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\syn}} \\
  &  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
  && \aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}
\end{flalign*}
With these operations, a straightforward check shows that every rule is mode-correct, therefore this as a mode-correct bidirectional type system.
As a result, we can instantiate the type synthesiser presented in \Cref{sec:type-synthesis} for this system, eliminating the need for re-implementation.

\subsection{Spine Calculus}\label{subsec:spine}
A spine application, denoted as $t\;u_1\;\ldots\;u_n$, is a variant of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
This arrangement allows direct access to the head term, making it practical in various applications.
For instance, \Agda's core language employs this form of application, as does its reflected syntax used for metaprogramming.

At first glance, accommodating this form of application may seem impossible, given that the number of arguments a construct can accept has to be fixed in a signature.
Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish corresponding constructs for each number $n$ of arguments, thereby exhibiting the necessity of having an arbitrary set (with decidable equality) for operation symbols.

### 2023-07-11 tex/sec4-pre-synthesis.tex
[ tex ] Revisions here and there

The goal of this sub-section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
In fact we will do better:
If a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
\footnote{For brevity, we use ordinary mode decoration elsewhere in this paper.}

@@

Intuitively, a term does not have a mode derivation exactly when there are not enough type annotations, but such negative formulations convey little information.
Instead, we can provide more information by pointing out the places in the term that require annotations.
For a bidirectional type system, an annotation is required wherever a term is `strictly' (which we will explain shortly) in checking mode but required to be in synthesising mode, in which case there is no rule for switching from checking to synthesising, and thus there is no way to construct a mode derivation.
We can, however, consider \emph{generalised mode derivations} defined in \cref{fig:generalised-mode-derivations} that allow the use of an additional $\SynRule{Missing}$ rule for such switching, so that a derivation can always be constructed.
Given a generalised mode derivation, if it uses $\SynRule{Missing}$ in some places, then those places are exactly where annotations should be supplied, which is helpful information to the user; if it does not use $\SynRule{Missing}$, then the derivation is \emph{genuine} in the sense that it corresponds directly to an original mode derivation.
This can be succinctly formulated as \cref{thm:Pre?-true} below by encoding genuineness as a boolean~$g$ in the generalised mode judgement, which is set to~$\BFalse$ only by the $\SynRule{Missing}$ rule.
(Ignore the boolean~$s$ for now.)

@@

\begin{proof}
Case analysis on $\dir{d}$ and $\dir{d'}$, adding an outermost mode cast to change the given derivation to a different mode if $d \neq d'$.
Note that it is permissible to add an outermost mode cast because the antecedent~(\ref{eq:strict-derivation}) requires the given derivation to be strict.
\end{proof}

@@

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be easily proved from generalised mode decoration.

>>>>

The major goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
In fact we will do better:
If a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
\footnote{For brevity, we use ordinary mode decoration elsewhere in this paper.}

@@

Intuitively, a term does not have a mode derivation exactly when there are not enough type annotations, but such negative formulations convey little information.
Instead, we can provide more information by pointing out the places in the term that require annotations.
For a bidirectional type system, an annotation is required wherever a term is `strictly' (which we will define shortly) in checking mode but required to be in synthesising mode, in which case there is no rule for switching from checking to synthesising, and thus there is no way to construct a mode derivation.
We can, however, consider \emph{generalised mode derivations} defined in \cref{fig:generalised-mode-derivations} that allow the use of an additional $\SynRule{Missing}$ rule for such switching, so that a derivation can always be constructed.
Given a generalised mode derivation, if it uses $\SynRule{Missing}$ in some places, then those places are exactly where annotations should be supplied, which is helpful information to the user; if it does not use $\SynRule{Missing}$, then the derivation is \emph{genuine} in the sense that it corresponds directly to an original mode derivation.
This can be succinctly formulated as \cref{thm:Pre?-true} below by encoding genuineness as a boolean~$g$ in the generalised mode judgement, which is set to~$\BFalse$ only by the $\SynRule{Missing}$ rule.
(Ignore the boolean~$s$ for now.)

@@

\begin{proof}
Case analysis on $\dir{d}$ and $\dir{d'}$, adding an outermost mode cast to change the given derivation to a different mode if $\dir{d} \neq \dir{d'}$.
Note that it is permissible to add an outermost mode cast because the antecedent~(\ref{eq:strict-derivation}) requires the given derivation to be strict.
\end{proof}

@@

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be easily proved from generalised mode decoration.
